---
title: "Spatial Topology Repair for Areal Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Topology Repair for Areal Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`## The Disconnected Graph Problem

Bayesian spatial disease mapping relies heavily on Conditional Autoregressive (CAR) priors, such as the BYM2 model implemented in `R-INLA`. These models assume that spatial dependence flows through a neighborhood matrix. However, an intrinsic requirement of these Gaussian Markov Random Fields (GMRF) is that the spatial graph must form a **single connected component** to satisfy strict sum-to-zero constraints.

When working at high resolutions (e.g., 100m grids), basic contiguity matrices frequently fracture. Water bodies, parks, and industrial zones split the grid into isolated islands and disconnected sub-graphs. 

The `nbstitch` package provides a programmatic toolkit to iteratively repair these topologies, utilizing infrastructure routing and spatial absorption to force convergence into a single component.

## Setup: A Fragmented Grid

Let's simulate a high-resolution urban environment. We will generate a spatial grid and intentionally remove cells to create a "river" splitting the city, along with an isolated "island" cell.

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(spdep)
library(nbstitch)

# 1. Generate a 10x10 projected spatial grid (UTM meters)
grid_sf <- st_make_grid(
  cellsize = 100, 
  offset = c(500000, 4500000), 
  n = c(10, 10)
) |> st_sf()
st_crs(grid_sf) <- 32618

# 2. Simulate geographical barriers (a river and an island)
# Remove a column to split the grid into Left and Right components
# Keep one isolated cell at the top right to act as an island
grid_sf$id <- 1:nrow(grid_sf)
grid_sf <- grid_sf[!(grid_sf$id %in% c(41:50, 99)), ]

# Extract coordinates and build the raw neighborhood graph
coord <- st_centroid(grid_sf)
raw_nb <- poly2nb(grid_sf, queen = TRUE)

# Check component count
g_raw <- igraph::graph_from_adjacency_matrix(nb2mat(raw_nb, style="B", zero.policy=TRUE), mode="undirected")
igraph::components(g_raw)$no
```

The raw graph is fractured into 3 distinct components. If we pass this to INLA, it will throw an improper GMRF error.

## The `nbstitch` Pipeline

We can repair this topology using the three-stage `stitch_nb()` pipeline. 

### Stage 1: Infrastructure Wormholes (`add_wormhole`)
Geographical gaps are often bridged by real-world infrastructure (subways, ferries, bridges). We can manually define these pathways to intelligently stitch major landmasses together.

```{r wormhole}
# Define a bridge crossing our simulated river
bridge_link <- data.frame(
  name = "Main_Bridge",
  from_x = 500350, from_y = 4500550,
  to_x   = 500550, to_y   = 4500550
)
```

### Stage 2 & 3: Absorption and Island Stitching (`sweep_absorp`, `stitch_island`)
Once primary infrastructure is defined, `nbstitch` iteratively locates any remaining orphan sub-graphs and isolated zero-neighbor nodes, snapping them to the nearest valid node in the giant component.

### Execution

Instead of running each step manually, we pass the data into the orchestrator function:

```{r execute}
# Execute the full repair pipeline
repaired_nb <- stitch_nb(
  nb = raw_nb,
  coord = coord,
  wormhole = bridge_link
)

# Verify graph health
summary(repaired_nb)
```

The graph has successfully converged into a single component! 

## Validation for INLA

Because `nbstitch` outputs a standard `nb` class object, the repaired topology is immediately ready for export to INLA's sparse matrix format.

```{r export, eval=FALSE}
# Write the repaired graph to a format INLA can ingest
nb2INLA("repaired_graph.adj", repaired_nb)
```
